\large{\bf Looking forwards and backwards: dynamics and genealogies of locally regulated populations }

### 2023-04-25 forwards-and-backwards.tex
done with printout pass through

Such results on (one-step) convergence to reaction diffusion 
equation limits have been
achieved for a variety of interacting particle systems. 
Following the now classical contributions 
of~\citet{demasi/ferrari/lebowitz:1986, demasi/presutti:1991, oeschlaeger:1985},
much of this work has focused on lattice based models with one particle
per site, or on systems with a fixed number, $N$,
of interacting diffusions as $N\to\infty$.
For systems of proliferating particles,
as considered for example by 
\citet{oelschlaeger:1989, flandoli/leimbach/olivera:2019, flandoli/huang:2021}, 
an additional challenge (also apparent in our models), is 
the control of concentration of particles. 
We follow 
\citet{oelschlaeger:1989, flandoli/leimbach/olivera:2019} in considering
`moderate interactions', meaning that
the number of individuals in the neighbourhood over which we measure local
population density tends to infinity, whereas 
\citet{flandoli/huang:2021} also consider the situation in which
that number remains finite. 
We refer to \citet{flandoli/huang:2021} for a more thorough 
literature review, 
but note that both our model and scaling differ from those considered
in the body of work discussed there: whereas in those settings, the only 
scalings are the number of particles in the system and the size of the
neighbourhood over which individuals interact with one another, in
keeping with the vast literature on continuous state branching models, we 
also scale time and so must ensure that births are 
adequately compensated by deaths to prevent the population from exploding.

@@

Theorem \ref{thm:nonlocal_convergence},
combined with Proposition \ref{prop:nonlocal_to_local} or~\label{nonlocalPME to PME}
implies that we can take the limit $N \to \infty$
followed by the limit $\epsilon \to 0$
to obtain solutions to the PDE \eqref{localPDE}.
However, it is of substantial interest to know whether
we can take those two limits simultaneously.
The general case seems difficult,
but we prove such ``diagonal'' convergence in the following situation.

@@

\begin{proof}[of Proposition~\ref{thm:mmt_application}]
    This proposition is the content of the Markov Mapping Theorem,
    reproduced from \citet{etheridge/kurtz:2019} as Theorem~\ref{thm:mmt},
    applied to our situation.
    In the notation of that theorem, $\gamma$ is what we have called $\kappa$ above,
    $\alpha$ is the transition function that assigns levels uniformly on $[0, N]$ (in the first case)
    or as a Poisson process with Lesbegue intensity (in the limiting case),
    and we need a continuous $\psi^N(\lp)$ such that $|A^N f(\lp)| \le c_f \psi^N(\lp)$
    for all $f$ in the domain of $A^N$ (and similarly a function $\psi$ for $A$).
    We also need that applying the lookdown generator to a function and averaging over levels
    is equivalent to applying the population process generator to the function
    whose dependence on levels has been averaged out,
    a condition which we precisely state, and verify,
    in Lemmas~\ref{lem:averaged_generators} and~\ref{lem:averaged_generators_limit}
    of the Appendix.

@@

\begin{remark}
The process we consider is similar to the state-dependent branching processes
of \citet{KR}, so one might expect that the proofs there would carry over with little change.
However, there is an important difference:
Recall that the level $U_a(t)$ of a line of descent evolves as
$$
\dot u = c_\theta(x, \eta) u^2 - b_\theta(x, \eta) u,
$$
where $b_\theta(x, \eta)$ and $c_\theta(x, \eta)$ are defined in \eqref{b_defn}
and \eqref{c_defn} respectively.
Note that $c_\theta(x, \eta) \ge 0$ and $b_\theta(x, \eta)$ may take either sign.
Assumptions~\ref{def:model_setup} imply that $c^N(x, \eta)$ is bounded,
while $b^N(x, \eta)$, because of $F(x, \eta)$, is bounded above but not necessarily below.
In \citet{kurtz/rodrigues:2011}, $b^N$ was bounded above and $c^N$ was bounded away from zero,
so they noted that if $U_a(t) \ge b^N/c^N$ for some label $a$,
that line of descent would only move upwards from that time onwards.
Furthermore, coefficients did not depend on the state of the process (i.e., on $\eta$),
thus allowing the processes to be jointly and simultaneously constructed
for all values of $N$, with a pointwise embedding
of $(\lp^N_t)_{t \ge 0}$ within $(\lp^{M})_{t \ge 0}$ for $b^N/c^N < N < M$.
In other words, individuals with levels above $N > b^N/c^N$ at time $t_0$
do not affect $(\lp^N_t)_{t \ge t_0}$,
thus allowing a comparison of the number of lines of descent below level $u_0$
to a branching process.
Although we have provided a
joint construction of $\lp^N$ for all $N$ in Section~\ref{sec: individual lines of descent},
it does not have this monotonicity:
for one thing, $b^N$ and $c^N$ depend on the population process $\eta$
and so all individuals can affect all other ones (even those with lower levels).
Furthermore, in the deterministic case $\theta/N$, and hence $c$, converges to zero,
and so lines of descent with arbitrarily high level may drift back downwards.
Indeed, this must be the case if the population persists,
since in the deterministic case there is no branching.
\end{remark}

To complete the description of the process,
each line of descent
gives birth to lines at higher levels at rate $2 (N - u) c_\theta(x, \eta)$,
and each such new line chooses a level uniformly from $[u, N]$,
a spatial location $y$ from the kernel
\begin{equation} \label{eqn:qm_defn}
    q^m(x, dy, \eta)
    =
    r(y, \eta) q(x, dy) / \int_{\IR^d} r(y, \eta) q(x, dy),
\end{equation}
and the two lines swap spatial locations with probability 1/2.

It is evident from the description of the process
(or, by differentiating in Definition~\ref{defn:lookdown_mgale})
that
\begin{align} \label{eqn:f_xi_mgale}
\begin{split}
    \langle f, \lp_t^N \rangle
    &=
    \langle f, \lp_0^N \rangle
    + M^f_t
\\ & \qquad {}
    +
    \int_0^t
    \big\langle
        c_\theta(x, \eta^N_s)
        \int_u^N \int_{\IR^d}
        \left(
            f(y, u_1) + f(x, u_1) + f(y, u) - f(x, u)
        \right)
        q^m(x, dy, \eta^N_s) du_1 
\\ & \qquad \qquad \qquad \qquad {}
        +
        \left(
            c_\theta(x, \eta^N_s) u^2
            - b_\theta(x, \eta^N_s) u
        \right)
        \frac{d}{du} f(x, u)
    ,
    \lp_s^N(dx, du) \bigg \rangle
    ds ,
\end{split}
\end{align}
where $M^f$ is a martingale
with quadratic variation
\begin{align} \label{eqn:f_xi_qv}
    \begin{split}
    [M^f]_t
    &=
    \int_0^t
    \bigg \langle c_\theta(x, \eta_s^N) \int_u^N \int_{\IR^d}
        \big[ f(y, u_1)^2
\\&\qquad \qquad {}
        + \left( f(x, u_1) + f(y, u) - f(x, u) \right)^2 \big]
    du_1 q^m(x, dy, \eta^N_s)
    ,
    \lp_s^N(dx, du) \bigg \rangle
    ds .
\end{split}\end{align}

In addition to tightness of the measure-values processes $\lp^N$,
the bounds used in the proofs below also imply
tightness of the number of lines of descent and the number of births below a fixed level,
and of the motion of individual lines of descent.
In other words, the limiting ``line of descent'' construction holds.

\begin{proof}[Proof of Theorem~\ref{thm:lookdown_convergence}]
    As in Section~\ref{sec:population_density_proof},
    the theorem will follow from tightness and characterization of the limit.
    This time, the processes $\lp^N$ take values in $\lpmeasures$,
    the space of locally finite measures on space $\times$ levels.
    (They will in fact be point measures, including the limit,
    but that is a consequence of this theorem.)
    Again, tightness follows from a compact containment condition,
    tightness of one-dimensional distributions,
    and an application of \citet{EK} Theorem~3.9.1.

@@

\begin{lemma}
        \label{lem:h_containment}
    Let $h$ be a positive, continuous, nonincreasing, differentiable function on $[0, \infty)$
    such that
    $\int_0^\infty \int_u^\infty h(v) dv du$,
    $\int_0^\infty u^2 |h'(u)| du$,
    and $\int_0^\infty h(u)^2 du$
    are all finite,
    Suppose that Assumptions~\ref{def:model_setup} hold,
    and that $\theta/N \to \alpha$ and $\lp_0^N \to \lp_0$ weakly as $N \to \infty$,
    where each $\lp_0^N$ is conditionally uniform given $\eta_0^N$
    in the sense of \eqref{eqn:conditionally_uniform}
    and $\lp_0$ is conditionally uniform given $\eta_0$
    in the sense of \eqref{eqn:conditionally_poisson}.
    Then for any $T$ there exists a constant $K(T)$ such that for all $M > 0$,
    $$ 
    \limsup_{N \to \infty} \IP\left\{ \sup_{0 \le t \le T} \langle h, \lp^N_t \rangle > M \right\}
    <
    \frac{K(T)}{M} .
    $$
\end{lemma}

@@

\begin{proof}[of Lemma~\ref{lem:h_containment}]
    Applied to $f(x, u) = h(u)$, the martinagle representation~\eqref{eqn:f_xi_mgale} is
    \begin{align*}
    \begin{split}
        \langle h, \lp_t^N \rangle
        &=
        \langle h, \lp_0^N \rangle
        + M^{h}_t 
        \\ & \qquad {}
        +
        \int_0^t
        \big\langle
            2 c^N(x, \eta^N_s) \int_u^N h(v) dv
        ,
        \lp_s^N(dx, du) \big \rangle
        ds
        \\ & \qquad \qquad {}
        +
        \int_0^t
        \big\langle
            \left(
                c^N(x, \eta^N_s) u^2
                - b^N(x, \eta^N_s) u
            \right)
            h'(u)
        ,
        \lp_s^N(dx, du) \big \rangle
        ds ,
    \end{split}
    \end{align*}
    where $M^{h}_t$ is a martingale
    with quadratic variation
    \begin{align*}
        [M^{h}]_t
        &=
        \int_0^t
        \langle 2 c^N(x, \eta_s^N) \int_u^N h(v)^2 dv
        ,
        \lp_s^N(dx, du) \big \rangle
        ds .
    \end{align*}
    Now, note that $0 \le c^N(x, \eta_x^N) \le C_a < \infty$
    and $b^N(x, \eta_s^N) \le C_b < \infty$,
    and we have assumed that $h'(u) \le 0$ (since $h$ is nonincreasing),
    so we may bound
    \begin{align} \label{eqn:hn_bound}
    \begin{split} 
        \langle h, \lp_t^N \rangle
        &\le
        \langle h, \lp_0^N \rangle
        + M^{h}_t
        \\ & \qquad {}
        +
        \int_0^t
        \big\langle
            2 C_a \int_u^\infty h(v) dv
            + \left( C_a u^2 + C_b u \right) |h'(u)|,
        \lp_s^N(dx, du) \big \rangle
        ds .
    \end{split}
    \end{align}
    Now, since $\lp_t^N$ is conditionally uniform given $\eta_t^N$,
    we know that for appropriate $f$,
    $  \IE[\langle f, \lp^N_t \rangle] = \IE[\langle \widetilde{f}, \eta^N_t \rangle] , $
    where $\widetilde{f}(x) = \int_0^\infty f(x, u) du$.
    By our assumptions on $h$,
    we know that
    $$
        \int_0^\infty \left( 2 C_a \int_u^\infty h(v) dv
        + \left( C_a u^2 + C_b u \right) |h'(u)| \right) du
        < C
    $$
    for some $C < \infty$, and so
    \begin{align*}
    \begin{split}
        \IE\left[ \langle h, \lp_t^N \rangle \right]
        &\le
        \IE\left[ \langle h, \lp_0^N \rangle \right]
        +
        C
        \int_0^t
        \IE\bigg[
            \big\langle
            1
            ,
            \eta_s^N \big \rangle
        \bigg]
        ds ,
    \end{split}
    \end{align*}
    which we know by Lemma~\ref{lem:eta_f_bound}
    is bounded by $C_0 e^{C_1 t}$ for some other constants $C_0$ and $C_1$.

@@

    Again since $\sqrt{x} \le 1 + x$ for $x \ge 0$,
    the Burkholder-Davis-Gundy inequality
    tells us that there is a $C'$ such that
    \begin{align*}
    \IE\left[ \sup_{0 \le t \le T} M^{h}_t \right]
    & \le
        C'\left( 1 + \IE\left[ [M^{h}]_T \right] \right)
    \\ & \le
        C' \left( 1 + \int_0^T \IE\left[
        \langle 2 c^N(x, \eta_s^N) \int_u^\infty h(v)^2 dv
        ,
        \lp_s^N(dx, du) \big \rangle
        \right] ds \right) 
    \\ & \le
        C' \left( 1 + 2 C_a \int_0^\infty h(v)^2 dv
        \int_0^T \IE\left[
            \langle 1 , \lp_s^N(dx, du) \big \rangle
        \right] ds \right)  
    \\ & \le
        C_2 e^{C_1 T} ,
    \end{align*}
    for a constant $C_2$ which is finite
    by our assumption that $\int_0^\infty h(v)^2 dv < \infty$.

@@

\begin{proof}[of Lemma \ref{lem:test_fn_tightness}]
    Again, we use the Aldous-Rebolledo criterion.
    Tightness of $\langle f, \lp_t \rangle$ for a fixed $t$
    follows from Lemma~\ref{lem:compact containment hn},
    so we need only prove conditions analogous to
    \eqref{eqn:eta_projections_goal1}
    and \eqref{eqn:eta_projections_goal2}
    applied to the martingale representation
    of equations~\eqref{eqn:f_xi_mgale} and~\eqref{eqn:f_xi_qv}.
    Rewriting~\eqref{eqn:f_xi_mgale} with $c_\theta = c_\theta(x, \eta_s)$,
    \begin{align*}
        \langle f, \lp_t \rangle
        &=
        \langle f, \lp_0 \rangle
        + M^f_t
        + \int_0^t \Big\langle
            c_\theta \int_u^N \int (f(y, u_1) + f(x, u_1)) q^m(x,dy,\eta) du_1
        \\ {} &\qquad
            + c_\theta (N - u) \int (f(y, u) - f(x, u)) q^m(x,dy,\eta)
            + (c_\theta u^2 - b_\theta u) \frac{d}{du} f,
            \lp_s
        \Big\rangle ds
        \\ &\le
        \langle f, \lp_0 \rangle
        + M^f_t
        + \int_0^t \Big\langle
            c_\theta \int_u^N \int (f(y, u_1) + f(x, u_1)) q^m(x,dy,\eta) du_1
        \\ {} &\qquad
            + c_\theta (N - u) \int (f(y, u) - f(x, u)) q^m(x,dy,\eta)
            + (c_\theta u^2 - b_\theta u) \frac{d}{du} f,
            \lp_s
        \Big\rangle ds
    \end{align*}
    The bounds analogous to \eqref{eqn:eta_projections_goal1} and \eqref{eqn:eta_projections_goal2}
    follow as in the proof of Lemma~\ref{lem:eta_projections_tightness}:
    for instance, observe that using that $c_\theta \le C_0$ for some $C_0$,
    the predictable part of this semimartingale decomposition is bounded by
    \begin{align*}
        \Big\langle
            2 C_0 \|f\|_\infty u_f
            + (1 - u/N) \gamma B^\theta_f
            + (C_0 u^2 - b_\theta u) \frac{d}{du} f,
            \lp_s
        \Big\rangle
        &\le
        \langle
            C_0 u_f^2 + \sup_x |b_\theta(x, \eta_s)| u_f \|\frac{d}{du} f\|_\infty
        \rangle ,
    \end{align*}
    which can be bounded as we did for \eqref{eqn:eta_projections_goal1}.
\end{proof}

@@

\begin{proof}[of Theorem~\ref{thm:lineages}]

Here we have taken the high-density, deterministic limit
(so, $\theta, N \to \infty$ and $\theta/N \to 0$)
and so suppose that the limiting process has density $\varphi_t(x)$
at location $x$ and time $t$ (which may only exist almost everywhere).
We first proceed informally, as if the density exists everywhere,
and follow this with an integration against test functions to make the argument rigorous.
Let $Y$ denote the spatial motion followed by a single line of descent.
Above equation~\eqref{eqn:limiting_generator},
we showed that $Y$ is a diffusion with generator at time $s$
$$
    \mathcal{L}^Y_s g(x) = \gamma ( \DG(r(\cdot,\eta_s) g(\cdot))(x) - g(x) \DG r(x,\eta_s) ) .
$$
% (This is $\mathcal{L}^Y_s g(x) = \gamma ( \DG(rg)(x) - g \DG r(x) )$ in shorter but more cryptic notation.)
The diffusion is time-inhomogeneous if the density is not constant with time
(since recall that the coefficients $\gamma$ and $r$ depend on the density).
Let $\varphi_t(x)$ be the limiting density, which solves \eqref{general deterministic limit},
$\partial_t \varphi = r \DG^*[ \varphi_t \gamma ] + \varphi F$.
Formally, the intensity of levels at $y$ at time $t$
that are descended from individuals that were at $x$ at time $s$
(with $s < t$) is
\begin{equation} \label{eqn:formal_intensity}
    \varphi_s(x) \IE_{s,x} \left[
        \exp\left(
            \int_s^t (F + \gamma \DG r)(Y_u) du
        \right)
        \ind_{Y_t = y}
    \right]
    dy ,
\end{equation}
where the subscript $s, x$ in the expectation indicates that $Y_s = x$.
To see why this should be true, 
suppose that an ancestor at time $s$ has level $v$. Conditional on its 
spatial motion $\{Y_u\}_{s\leq u\leq t}$, its level at time $t$ will
be $v \exp(-\int_s^t(F+\gamma\Delta r)(Y_u)du)$. This will be less than a given level 
$\lambda$ if $v < \lambda \exp(\int_s^t(F+\gamma\Delta r)(Y_u)du)$. 
The intensity of levels at $y$ that are descended from individuals at
$x$ can therefore be obtained as the limit as $\lambda\to\infty$ of 
$1/\lambda$ times the number of levels at $x$ at time zero with
$u<\lambda \exp(\int_s^t(F+\gamma\Delta r)(Y_u)du)$ and for which
the corresponding individual is at $y$ at time $t$, which is 
precisely the quantity in~\eqref{eqn:formal_intensity}.

>>>>

Such results on (one-step) convergence to reaction diffusion 
equation limits have been
achieved for a variety of interacting particle systems. 
Following the now classical contributions 
of~\citet{demasi/ferrari/lebowitz:1986, demasi/presutti:1991, oelschlaeger:1985},
much of this work has focused on lattice based models with one particle
per site, or on systems with a fixed number, $N$,
of interacting diffusions as $N\to\infty$.
For systems of proliferating particles,
as considered for example by 
\citet{oelschlaeger:1989, flandoli/leimbach/olivera:2019, flandoli/huang:2021}, 
an additional challenge (also apparent in our models), is 
the control of concentration of particles. 
We follow 
\citet{oelschlaeger:1989, flandoli/leimbach/olivera:2019} in considering
`moderate interactions', meaning that
the number of individuals in the neighbourhood over which we measure local
population density tends to infinity, whereas 
\citet{flandoli/huang:2021} also consider the situation in which
that number remains finite. 
We refer to \citet{flandoli/huang:2021} for a more thorough 
literature review, 
but note that both our model and scaling differ from those considered
in the body of work discussed there: whereas in those settings, the only 
scalings are the number of particles in the system and the size of the
neighbourhood over which individuals interact with one another, in
keeping with the vast literature on continuous state branching models, we 
also scale time and so must ensure that births are 
adequately compensated by deaths to prevent the population from exploding.

@@

Theorem \ref{thm:nonlocal_convergence},
combined with Proposition \ref{prop:nonlocal_to_local} or~\ref{nonlocalPME to PME}
implies that we can take the limit $N \to \infty$
followed by the limit $\epsilon \to 0$
to obtain solutions to the PDE \eqref{localPDE}.
However, it is of substantial interest to know whether
we can take those two limits simultaneously.
The general case seems difficult,
but we prove such ``diagonal'' convergence in the following situation.

@@

\begin{proof}[of Proposition~\ref{thm:mmt_application}]
    This proposition is the content of the Markov Mapping Theorem,
    reproduced from \citet{etheridge/kurtz:2019} as Theorem~\ref{thm:mmt},
    applied to our situation.
    The function $\gamma$ of that theorem is what we have called $\kappa$ above,
    and the kernel $\alpha$ of that theorem
    is the transition function that assigns levels uniformly on $[0, N]$ (in the first case)
    or as a Poisson process with Lesbegue intensity (in the limiting case).
    We need a continuous $\psi^N(\lp)$ such that $|A^N f(\lp)| \le c_f \psi^N(\lp)$
    for all $f$ in the domain of $A^N$ (and similarly a function $\psi$ for $A$).
    We also need that applying the lookdown generator to a function and averaging over levels
    is equivalent to applying the population process generator to the function
    whose dependence on levels has been averaged out,
    a condition which we precisely state, and verify,
    in Lemmas~\ref{lem:averaged_generators} and~\ref{lem:averaged_generators_limit}
    of the Appendix.

@@

\begin{remark}
The process we consider is similar to the state-dependent branching processes
of \citet{kurtz/rodrigues:2011}, so one might expect that the proofs there would carry over with little change.
However, there is an important difference:
Recall that the level $U_a(t)$ of a line of descent evolves as
\begin{equation} \label{eqn:u_evolution}
\dot u = c_\theta(x, \eta) u^2 - b_\theta(x, \eta) u,
\end{equation}
where $b_\theta(x, \eta)$ and $c_\theta(x, \eta)$ are defined in \eqref{b_defn}
and \eqref{c_defn} respectively.
Note that $c_\theta(x, \eta) \ge 0$, while $b_\theta(x, \eta)$ may take either sign.
Assumptions~\ref{def:model_setup} imply that $c_\theta(x, \eta)$ is bounded,
while $b_\theta(x, \eta)$, because of $F(x, \eta)$, is bounded above but not necessarily below.
In \citet{kurtz/rodrigues:2011}, $b_\theta$ was bounded above and $c_\theta$ was bounded away from zero,
so they noted that if $U_a(t) \ge b_\theta/c_\theta$ for some label $a$,
that line of descent would only move upwards from that time onwards.
Furthermore, coefficients did not depend on the state of the process (i.e., on $\eta$),
thus allowing the processes to be jointly and simultaneously constructed
for all values of $N$, with a pointwise embedding
of $(\lp^N_t)_{t \ge 0}$ within $(\lp^{M})_{t \ge 0}$ for $b_\theta/c_\theta < N < M$.
In other words, individuals with levels above $N > b_\theta/c_\theta$ at time $t_0$
do not affect $(\lp^N_t)_{t \ge t_0}$,
thus allowing a comparison of the number of lines of descent below level $u_0$
to a branching process.
Although we have provided a
joint construction of $\lp^N$ for all $N$ in Section~\ref{sec: individual lines of descent},
it does not have this monotonicity:
for one thing, $b_\theta$ and $c_\theta$ depend on the population process $\eta$
and so all individuals can affect all other ones (even those with lower levels).
Furthermore, in the deterministic case $\theta/N$, and hence $c$, converges to zero,
and so lines of descent with arbitrarily high level may drift back downwards.
Indeed, this must be the case if the population persists,
since in the deterministic case there is no branching.
\end{remark}

We first give a description of the lookdown process $\lp^N$ in terms of the lines of descent
introduced in Section~\ref{sec: individual lines of descent}.
Each line of descent
gives birth to lines at higher levels at rate $2 (N - u) c_\theta(x, \eta)$,
and each such new line chooses a level uniformly from $[u, N]$,
a spatial location $y$ from the kernel
\begin{equation} \label{eqn:qm_defn}
    q^m(x, dy, \eta)
    =
    r(y, \eta) q(x, dy) / \int_{\IR^d} r(y, \eta) q(x, dy),
\end{equation}
and the two lines swap spatial locations with probability 1/2;
the level of each line of descent evolves according to equation~\eqref{eqn:u_evolution}.

It is evident from the description of the process
(or, by differentiating in Definition~\ref{defn:lookdown_mgale})
that
\begin{align} \label{eqn:f_xi_mgale}
\begin{split}
    \langle f, \lp_t^N \rangle
    &=
    \langle f, \lp_0^N \rangle
    + M^f_t
\\ & \qquad {}
    +
    \int_0^t
    \bigg\langle
        c_\theta(x, \eta^N_s)
        \int_u^N \int_{\IR^d}
        \left(
            f(y, u_1) + f(x, u_1) + f(y, u) - f(x, u)
        \right)
        q^m(x, dy, \eta^N_s) du_1 
\\ & \qquad \qquad \qquad \qquad {}
        +
        \left(
            c_\theta(x, \eta^N_s) u^2
            - b_\theta(x, \eta^N_s) u
        \right)
        \frac{d}{du} f(x, u)
    ,
    \lp_s^N(dx, du) \bigg \rangle
    ds ,
\end{split}
\end{align}
where $M^f$ is a martingale
with quadratic variation
\begin{align} \label{eqn:f_xi_qv}
    \begin{split}
    [M^f]_t
    &=
    \int_0^t
    \bigg \langle c_\theta(x, \eta_s^N) \int_u^N \int_{\IR^d}
        \big[ f(y, u_1)^2
\\&\qquad \qquad {}
        + \left( f(x, u_1) + f(y, u) - f(x, u) \right)^2 \big]
    du_1 q^m(x, dy, \eta^N_s)
    ,
    \lp_s^N(dx, du) \bigg \rangle
    ds .
\end{split}\end{align}

In addition to tightness of the measure-values processes $\lp^N$,
the bounds used in the proofs below also imply
tightness of the number of lines of descent and the number of births below a fixed level,
and of the motion of individual lines of descent.
In other words, the limiting ``line of descent'' construction
of Section~\ref{sec: individual lines of descent} holds.

\begin{proof}[Proof of Theorem~\ref{thm:lookdown_convergence}]
    As in Section~\ref{sec:population_density_proof},
    the theorem will follow from tightness and characterization of the limit.
    This time, the processes $\lp^N$ take values in $\lpmeasures$,
    the space of locally finite measures on space $\times$ levels.
    (They will in fact be point measures, including the limit,
    but that is a consequence of this theorem.)
    Again, tightness follows from a compact containment condition,
    tightness of one-dimensional distributions,
    and an application of \citet{ethier/kurtz:1986} Theorem~3.9.1.

@@

\begin{lemma}
        \label{lem:h_containment}
    Let $h$ be a positive, continuous, nonincreasing, differentiable function on $[0, \infty)$
    such that
    $\int_0^\infty \int_u^\infty h(v) dv du$,
    $\int_0^\infty u^2 |h'(u)| du$,
    and $\int_0^\infty h(u)^2 du$
    are all finite,
    Suppose that Assumptions~\ref{def:model_setup} hold,
    and that $\theta/N \to \alpha$ and $\lp_0^N \to \lp_0$ weakly as $N \to \infty$,
    where each $\lp_0^N$ is conditionally uniform given $\eta_0^N$
    in the sense of \eqref{eqn:conditionally_uniform}
    and $\lp_0$ is conditionally Poisson given $\eta_0$
    in the sense of \eqref{eqn:conditionally_poisson}.
    Then for any $T$ there exists a constant $K(T)$ such that for all $M > 0$,
    $$ 
    \limsup_{N \to \infty} \IP\left\{ \sup_{0 \le t \le T} \langle h, \lp^N_t \rangle > M \right\}
    <
    \frac{K(T)}{M} .
    $$
\end{lemma}

@@

\begin{proof}[of Lemma~\ref{lem:h_containment}]
    Applied to $f(x, u) = h(u)$, the martinagle representation~\eqref{eqn:f_xi_mgale} is
    \begin{align*}
    \begin{split}
        \langle h, \lp_t^N \rangle
        &=
        \langle h, \lp_0^N \rangle
        + M^{h}_t 
        \\ & \qquad {}
        +
        \int_0^t
        \big\langle
            2 c_\theta(x, \eta^N_s) \int_u^N h(v) dv
        ,
        \lp_s^N(dx, du) \big \rangle
        ds
        \\ & \qquad \qquad {}
        +
        \int_0^t
        \big\langle
            \left(
                c_\theta(x, \eta^N_s) u^2
                - b_\theta(x, \eta^N_s) u
            \right)
            h'(u)
        ,
        \lp_s^N(dx, du) \big \rangle
        ds ,
    \end{split}
    \end{align*}
    where $M^{h}_t$ is a martingale
    with quadratic variation
    \begin{align*}
        [M^{h}]_t
        &=
        \int_0^t
        \langle 2 c_\theta(x, \eta_s^N) \int_u^N h(v)^2 dv
        ,
        \lp_s^N(dx, du) \big \rangle
        ds .
    \end{align*}
    Now, note that $0 \le c_\theta(x, \eta_x^N) \le C_a < \infty$
    and $b_\theta(x, \eta_s^N) \le C_b < \infty$,
    and we have assumed that $h'(u) \le 0$ (since $h$ is nonincreasing),
    so we may bound
    \begin{align} \label{eqn:hn_bound}
    \begin{split} 
        \langle h, \lp_t^N \rangle
        &\le
        \langle h, \lp_0^N \rangle
        + M^{h}_t
        \\ & \qquad {}
        +
        \int_0^t
        \big\langle
            2 C_a \int_u^\infty h(v) dv
            + \left( C_a u^2 + C_b u \right) |h'(u)|,
        \lp_s^N(dx, du) \big \rangle
        ds .
    \end{split}
    \end{align}
    Now, since $\lp_t^N$ is conditionally uniform given $\eta_t^N$,
    we know that for appropriate $f$,
    $  \IE[\langle f, \lp^N_t \rangle] = \IE[\langle \widetilde{f}, \eta^N_t \rangle] , $
    where $\widetilde{f}(x) = \int_0^\infty f(x, u) du$.
    By our assumptions on $h$,
    we know that
    $$
        \int_0^\infty \left( 2 C_a \int_u^\infty h(v) dv
        + \left( C_a u^2 + C_b u \right) |h'(u)| \right) du
        < C
    $$
    for some $C < \infty$, and so
    \begin{align*}
    \begin{split}
        \IE\left[ \langle h, \lp_t^N \rangle \right]
        &\le
        \IE\left[ \langle h, \lp_0^N \rangle \right]
        +
        C
        \int_0^t
        \IE\bigg[
            \big\langle
            1
            ,
            \eta_s^N \big \rangle
        \bigg]
        ds ,
    \end{split}
    \end{align*}
    which we know by Lemma~\ref{lem:eta_f_bound}
    is bounded by $C_0 e^{C_1 t}$ for some other constants $C_0$ and $C_1$.

@@

    Again since $\sqrt{x} \le 1 + x$ for $x \ge 0$,
    the Burkholder-Davis-Gundy inequality
    tells us that there is a $C'$ such that
    \begin{align*}
    \IE\left[ \sup_{0 \le t \le T} M^{h}_t \right]
    & \le
        C'\left( 1 + \IE\left[ [M^{h}]_T \right] \right)
    \\ & \le
        C' \left( 1 + \int_0^T \IE\left[
        \langle 2 c_\theta(x, \eta_s^N) \int_u^\infty h(v)^2 dv
        ,
        \lp_s^N(dx, du) \big \rangle
        \right] ds \right) 
    \\ & \le
        C' \left( 1 + 2 C_a \int_0^\infty h(v)^2 dv
        \int_0^T \IE\left[
            \langle 1 , \lp_s^N(dx, du) \big \rangle
        \right] ds \right)  
    \\ & \le
        C_2 e^{C_1 T} ,
    \end{align*}
    for a constant $C_2$ which is finite
    by our assumption that $\int_0^\infty h(v)^2 dv < \infty$.

@@

\begin{proof}[of Lemma \ref{lem:test_fn_tightness}]
    Again, we use the Aldous-Rebolledo criterion.
    Tightness of $\langle f, \lp_t \rangle$ for a fixed $t$
    follows from Lemma~\ref{lem:compact containment hn},
    so we need only prove conditions analogous to
    \eqref{eqn:eta_projections_goal1}
    and \eqref{eqn:eta_projections_goal2}
    applied to the martingale representation
    of equations~\eqref{eqn:f_xi_mgale} and~\eqref{eqn:f_xi_qv}.
    Rewriting~\eqref{eqn:f_xi_mgale} with $c_\theta = c_\theta(x, \eta_s)$,
    \begin{align*}
        \langle f, \lp_t \rangle
        &=
        \langle f, \lp_0 \rangle
        + M^f_t
        + \int_0^t \Big\langle
            c_\theta \int_u^N \int (f(y, u_1) + f(x, u_1)) q^m(x,dy,\eta) du_1
        \\ {} &\qquad
            + c_\theta (N - u) \int (f(y, u) - f(x, u)) q^m(x,dy,\eta)
            + (c_\theta u^2 - b_\theta u) \frac{d}{du} f,
            \lp_s
        \Big\rangle ds
    \end{align*}
    The bounds analogous to \eqref{eqn:eta_projections_goal1} and \eqref{eqn:eta_projections_goal2}
    follow as in the proof of Lemma~\ref{lem:eta_projections_tightness}:
    for instance, observe that using that $c_\theta \le C_a$ for some $C_a$,
    the predictable part of this semimartingale decomposition is bounded by
    \begin{align*}
        \Big\langle
            2 C_a \|f\|_\infty u_f
            + (1 - u/N) \gamma B^\theta_f
            + (C_a u^2 - b_\theta u) \frac{d}{du} f,
            \lp_s
        \Big\rangle ,
    \end{align*}
    the last term of which is bounded by
    \begin{align*}
        \langle
            C_a u_f^2 + \sup_x |b_\theta(x, \eta_s)| u_f \|\frac{d}{du} f\|_\infty
        \rangle ,
    \end{align*}
    which can be bounded as we did for \eqref{eqn:eta_projections_goal1}.
\end{proof}

@@

\begin{proof}[of Theorem~\ref{thm:lineages}]
For brevity, in the proof we write $\gamma(x)$ or $\gamma$ for $\gamma(x,\eta)$.

Here we have taken the high-density, deterministic limit
(so, $\theta, N \to \infty$ and $\theta/N \to 0$)
and so suppose that the limiting process has density $\varphi_t(x)$
at location $x$ and time $t$ (which may only exist almost everywhere).
We first proceed informally, as if the density exists everywhere,
and follow this with an integration against test functions to make the argument rigorous.
Let $Y$ denote the spatial motion followed by a single line of descent.
Above equation~\eqref{eqn:limiting_generator},
we showed that $Y$ is a diffusion with generator at time $s$
$$
    \mathcal{L}^Y_s g(x) = \gamma(x,\eta_s) ( \DG(r(\cdot,\eta_s) g(\cdot))(x) - g(x) \DG r(x,\eta_s) ) .
$$
% (This is $\mathcal{L}^Y_s g(x) = \gamma ( \DG(rg)(x) - g \DG r(x) )$ in shorter but more cryptic notation.)
The diffusion is time-inhomogeneous if the density is not constant with time.
Let $\varphi_t(x)$ be the limiting density, which solves \eqref{general deterministic limit},
$\partial_t \varphi = r \DG^*[ \varphi_t \gamma ] + \varphi F$.
Formally, the intensity of individuals at $y$ at time $t$
that are descended from individuals that were at $x$ at time $s$
(with $s < t$) is
\begin{equation} \label{eqn:formal_intensity}
    \varphi_s(x) \IE_{s,x} \left[
        \exp\left(
            \int_s^t (F + \gamma \DG r)(Y_u) du
        \right)
        \ind_{Y_t = y}
    \right]
    dy ,
\end{equation}
where the subscript $s, x$ in the expectation indicates that $Y_s = x$.
To see why this should be true, 
suppose that an ancestor at time $s$ has level $v$. Conditional on its 
spatial motion $\{Y_u\}_{s\leq u\leq t}$, its level at time $t$ will
be $v \exp(-\int_s^t(F+\gamma\DG r)(Y_u)du)$. This will be less than a given level 
$\lambda$ if $v < \lambda \exp(\int_s^t(F+\gamma\DG r)(Y_u)du)$. 
The intensity of levels at $y$ that are descended from individuals at
$x$ can therefore be obtained as the limit as $\lambda\to\infty$ of 
$1/\lambda$ times the number of levels at $x$ at time $s$ with
$u<\lambda \exp(\int_s^t(F+\gamma\DG r)(Y_u)du)$ and for which
the corresponding individual is at $y$ at time $t$, which is 
precisely the quantity in~\eqref{eqn:formal_intensity}.
